---
title: "Hands-on Exercise 1"
execute: 
  warning: false
date: "`r Sys.Date()`"
---

# Getting Started

## Installing and Loading R packages

In this section, I will install and load tidyverse and sf packages.

```{r}
pacman::p_load(tidyverse, sf)
```

## Importing a polygon feature data in shapefile format.

### Importing MasterPlan data

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_WEB_PL")
```

### Importing Cycling Path data

```{r}
cyclingpath <- st_read(dsn = "data/geospatial",
                layer = "CyclingPathGazette"
)
```

### Importing GIS data in kml format

```{r}
preschool <- st_read(dsn="data/geospatial/PreSchoolsLocation.kml")
```


## Plotting the geospatial data

```{r}
plot(mpsz)

plot(st_geometry(mpsz))

plot(mpsz["PLN_AREA_N"])

```

Correcting the coordinate system
```{r}

st_crs(mpsz)

mpsz3414 <- st_set_crs(mpsz, 3414)

st_crs(mpsz3414)

```
## Transforming the projection of preschool from wgs84 to svy21.

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

## Importing Aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")

list(listings) 

listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

## Buffering

```{r}

buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
buffer_cycling$AREA <- st_area(buffer_cycling)
sum(buffer_cycling$AREA)
```
